Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-10-26 manu-draft.qmd
more methods

Two primary sources of data have been used to track seagrass change over time in Tampa Bay. The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s \[e.g., , @SWFWMD23\]. These maps are created by photointerpretation of aerial images obtained at the end of the growing season, typically during November-December. The TBEP has used these maps to track progress towards achieving seagrass restoration goals as total cover in Tampa Bay. No species information is provided. A more detailed, but spatially-specific, data source is the Tampa Bay Interagency Seagrass Monitoring Program (Figure \[-@fig-seagrasschg\]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>). Annual transect surveys have been conducted since 1998 at 61 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017]. This dataset provides species information, including abundance, cover, frequency occurrence, and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed. Although the areal maps provide the standard for assessment of restoration goals, the transect data allow for inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined. As such, the transect data were used below for comparison with temperature and salinity changes for the major bay segments. Other sources of seagrass data are described in the next section.

@@

All of the organizations that provided water quality datasets participate in the Southwest Florida Regional Ambient Monitoring Program. This *ad hoc* group has routinely met quarterly to ensure similar standards and methods are used for the collection and processing of surface water quality monitoring data. Split-samples between the organizations are also compared to assess precision between different laboratories. As such, the water quality measurements used herein are considered comparable, relative to the different sampling designs used by each program.

@@

The first goal of this analysis was to describe spatial and temporal trends in water temperature and salinity using the three water quality datasets described above. This assessment provided an indication of the extent of change in Tampa Bay as context for understanding potential relationships with seagrass change. An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions from long-term climate change. Meteorological data describing air temperature and precipitation were obtained from Tampa International Airport where daily measurements have been collected since 1939. The *rnoaa* R package [@chamberlain23] was used to retrieve these data starting from 1975 when EPC monitoring began in Tampa Bay. Annual averages for air temperature and cumulative annual precipitation were calculated. Additionally, the Standardized Precipitation Index (SPI) was estimated from the daily rainfall data to identify periods of time when rainfall significantly deviated from the long-term average (using the spi function from...).

@@

Observed trends aggregated by year and bay segment across stations and months in @fig-meteowqraw.

The conceptual stressor diagram @fig-concept.

Salinity tolerance: Ruppia \> Halodule\> Thalassia \> Syringodium

Halodule:

-   Temp: 20-30 (range, @lewis1985)

-   Salinity: grows well in most salinity ranges [@lirman2003], like Syringodium but more tolerant [@lewis1985], 10-35 (range) or \< 10 (S. Scolaro)

Thalassia:

-   Temp: 20-30 (range, @lewis1985, @zieman1975), 30-31 (upper, from SS)

-   Salinity: 25 (lower, @lewis1985); 30, lowest at 5 (lower, @lirman2003), 33-38 (range, @phillips1960); 24-35 (range, @zieman1975 citing @phillips1960); 15 (SS)

Syringodium:

-   Temp: 20-30 (range, @lewis1985), 29 (upper from SS)

-   Salinity: 20 (lower, @lewis1985), 25 (lower, @lirman2003); 20

-   Get FIM, Pinellas, Manatee data

@@

>>>>

Two primary sources of data have been used to track seagrass change over time in Tampa Bay. The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s \[e.g., @SWFWMD23\]. These maps are created by photointerpretation of aerial images obtained at the end of the growing season, typically during November-December. The TBEP has used these maps to track progress towards achieving seagrass restoration goals as total cover in Tampa Bay. No species information is provided. A more detailed, but spatially-specific, data source is the Tampa Bay Interagency Seagrass Monitoring Program (Figure \[-@fig-seagrasschg\]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>). Annual transect surveys have been conducted since 1998 at 61 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017]. This dataset provides species information, including abundance, cover, frequency occurrence, and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed. Although the areal maps provide the standard for assessment of restoration goals, the transect data allow for inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined. As such, the transect data were used below for comparison with temperature and salinity changes for the major bay segments. Other sources of seagrass data are described in the next section.

@@

All of the organizations that provided water quality datasets participate in the Southwest Florida Regional Ambient Monitoring Program. This *ad hoc* group has routinely met quarterly to ensure similar standards and methods are used for the collection and processing of surface water quality monitoring data. Split-samples evaluated by each organization are also compared to assess precision between different laboratories. As such, the water quality measurements used herein are considered comparable, relative to the different sampling designs used by each program.

@@

The first goal of the analysis was to describe spatial and temporal trends in water temperature and salinity using the three water quality datasets described above. This assessment provided an indication of the extent of change in Tampa Bay as context for understanding potential relationships with seagrass change. An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change. Meteorological data describing air temperature and precipitation were obtained from Tampa International Airport where daily measurements have been collected since 1939. The *rnoaa* R package [@chamberlain23] was used to retrieve these data starting from 1975 when EPC monitoring began in Tampa Bay. Annual averages for air temperature and cumulative annual precipitation were calculated. Additionally, the Standardized Precipitation Index [SPI, @begueria2013] was estimated from the daily rainfall data to identify periods of time when rainfall significantly deviated from the long-term average [using the *spei* R package, @begueria23]. Annual hydrologic loading data to Tampa Bay beginning in 1985 were also obtained for comparison to annual precipitation [@tbep0623]. All climate and loading data were evaluated annually with simple linear regression trends to assess change over time. Water temperature and salinity trends using the EPC, FIM, and PDEM data were similarly evaluated by averaging the monthly data by year for each bay segment.

Formal trend tests were used to assess station-level changes in water temperature and salinity in the EPC data. These analyses also provided a detailed spatial assessment of trends because the EPC data is the only dataset of the three where the same sites have been sampled over time. Seasonal Kendall trend tests were used to evaluate the monotonic change for temperature and salinity from 1975 to present at each water quality station [@hirsch1982; @millard13]. The change per year was also evaluated for each parameter based on the slope estimates returned by each test. Kendall tests were also used to evaluate changes over time for each month to determine when the trends more most pronounced seasonally, e.g., all January estimates across years, all February estimates, etc. The percentage of stations in each bay segment with significant increasing temperature or decreasing salinity trends were evaluated for each month. All tests evaluated both surface and bottom measurements to assess potential differences by water depth.

@@

The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temparerature and salinity, with particular attention on the time periods before and after 2016 (pre/post recovery). The conceptual model for evaluating these changes follows simple niche space theory where seagrass growth and reproduction is hypothesized to be greatest within optimal ranges for forcing factors that are present in the environment [@hutchinson1957; @vandermeer1972]. In the simplest form, this can conceptualized as a bell curve with optimal conditions defined with a range of values (e.g., minimum and maximum temperatures where a species is typically observed), where reduced growth or mortality is observed outside of these ranges. Because both water temperature and salinity are evaluated herein, the same model can be conceptualized in two-dimensional space [@fig-concept]. Seagrass growth can be limited when temperature is below or above the optimum range, when salinity is below or above the optimum, or when both temperature and salinity conditions are outside of the optimum range. Using this conceptual model, we can evaluate whether seagrasses are individually stressed by either temperature or salinity, or whether seagrasses are jointly stressed by both temperature and salinity. Based on the results of the trend tests, we hypothesized that seagrasses are likely stressed by both high temperature and low salinity. The statistical models described below were constructed specifically to assess these hypotheses following the conceptual model in @fig-concept. Of course, the optimal niche space can be defined in n-dimensions, but we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have generally not been limiting for growth in recent years.

A fundamental challenge in describing niche space is identifying the boundaries within which the space occurs. In reality, the boundaries are typically not discrete and organisms can be found outside of these ranges, although at reduced abundances. Regardless, discrete limits are often used and substantial efforts have been made to identify these boundaries for seagrasses. In Tampa Bay, three dominant seagrass species are observed that include *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinium* (turtle grass) [@lewis1985]. These species co-occur often in mixed beds, although differences in abundance are observed across salinity ranges. Shoal grass is tolerant of a wide range of salinity [@lirman2003], but is more abundant in oligohaline portions of Tampa Bay. Conversely, turtle grass is most often found in more euryhaline conditions near the mouth of Tampa Bay. Reported salinity ranges for each of these species varies depending on location, season, and other co-occurring factors like temperature [@phillips1960; @mcmillan1967; @zieman1975; @lewis1985], although most studies place lower limits of salinity in the range of 15-25 ppt. Optimal temperature ranges are similar between species, with reduced growth observed at temperatures above 30$^\circ$ C [@zieman1975;@lewis1985].

Because of the uncertainty in defining thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to both describe the potential for stress and how it may be related to changes in seagrass. Distinctions were not made between species, primarily due to lack of consensus between studies and the likely site-specific ranges that affect seagrass growth in Tampa Bay. First, we developed metrics of potential temperature and salinity stress by quantifying the maximum number of continuous days each year when temperature was above or salinity was below a given threshold. This approach assumed that stress could be observed based on duration of exposure (i.e., maximum number of continuous days each year) relative to a threshold that may are may not be outside of the optimum range for seagrasses. These metrics were quantified from the monthly long-term observation in the EPC data. To quantify daily counts each year, a continuous prediction of temperature or stress over time was estimated using Generalized Additive Models (GAMs) fit to the response variable and a single predictor for decimal year. The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends [@beck2022a]. Model fit for each station was sufficient to calculate daily predictions for assessing the stressor metrics.

Counts of the maximum number of days each year that temperature was above or salinity was below a thresold were obtained from the GAMs. This was done at each of 45 stations in the EPC data using temperature thresholds of 29, 30, and 31$^\circ$ C and salinity thresholds of 15, 20, and 25 ppt.

@@

Role of additional stressors and why they weren't considered in the model - they're important but we can also demonstrate that light attenuation has been relatively stable over time, so may not be important to include in the models.

Issues with thresholds, not definitive and those likely to stress seagrass are potentially not occurring at frequencies sufficient enough to model. Those used herein were chosen statistically. Point is that the temp and sal trends are in a direction such that thresholds that are more likely to stress seagrass are more likely to occur in the future.

Case for continuous monitoring

Are we just showing seagrasses are not present because of temperature limitations or if temperature indeed caused a reduction?

Manamagement implication - we can't do much to control temp and salinity, but we can temper expectations for how to manage the bay. Adaptive capacity/resiliency is likely decreasing because of these "co-morbidities", so we have to temper expectations for other stressors that we can control.

OTB is likely at a tipping point...
